---
title: "Working with RangedSummarizedExperiments with DNA Methylation Data"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{identifying_tmrs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 6.75
)
```

```{r setup, message = F}
library(methodical)
library(TCGAbiolinks)
```

## Introduction

Most functions from `methodical` take as input RangedSumamrizedExperiment objects 
with methylation data. If there are many samples, there can be many millions
or, in the case of WGBS data, even billions of data points within a DNA methylation 
dataset. It can be unfeasible to load all this data into memory at once. This problem 
can be overcome by using DelayedArrays backed by HDF5 files, enabling data to be 
read into memory only as needed. Methodical provides a suite of functions 
for working with such DNA methylation RangedSumamrizedExperiments, including 
functions to extract methylation values for sites overlapping genomic regions of 
interest `GRanges`, to randomly sample methylation sites, to liftover the
methylation sites from one genome build to another and to mask methylation sites
overlapping certain genomic regions, e.g. repeats. 

Here we demonstrate this different functionality using a dataset downloaded
from TumourMethData.

tcga_wgbs_hg38 = HDF5Array::loadHDF5SummarizedExperiment("~/wgbs/tcga_wgbs/tcga_wgbs_hg38/")

## Extracting methylation data from methylation RangedSummarizedExperiments

We'll first demonstrate how to extract methylation data for CpG islands from
tcga_wgbs_hg38. 

```{r, eval=TRUE}
# Load CpG islands annotation for hg38
cpg_islands_hg38 <- annotatr::build_annotations(genome = "hg38", annotations = "hg38_cpgs")

```

## Creation of methylation RSE from array files

Generally, files from methylation array files will have one column with the 
name of the probe used to measure methylation at a particular CpG site and
another column with the value associated with this probe. 
`makeMethRSEFromArrayFiles` takes a GRanges object for the `probe_ranges` argument 
to match the names of these probe to genomic coordinates when creating the meth RSE. 
This should be a GRanges object with one metadata object called "name" which 
gives the names of the probes and matches the names in the array files. 

One such GRanges object, `infinium_450k_probe_granges_hg19` for mapping probes 
from the Illumina HumanMethylation450 array to genomic coordinates of the hg19 
human genome build is provided as part of the package. If mapping the probes to 
a genome build besides hg19 or using a different methylation array, another 
similar GRanges object with the coordinates of probes and a metadata column
with their names should be provided.

We'll take a look at one of the methylation files we just downloaded and at
`infinium_450k_probe_granges_hg19`.

```{r, eval=TRUE}
# View the top of one of the COAD methylation array files
system2("head", meth_array_files[1])

# Load infinium_450k_probe_granges_hg19 and view the first few ranges
data("infinium_450k_probe_granges_hg19")
head(infinium_450k_probe_granges_hg19)
```

Now we'll create the meth RSE for the four files we downloaded. It is assumed
that the first column in the methylation array files is the name of the probes
and the second is the values associated with them but this can be changed using
the `probe_name_column` and `beta_value_column` arguments. 

Files can be processed in parallel using BiocParallel with the `BPPARAM` argument,
which defaults to `BiocParallel::bpparam()`. 

We'll save the HDF5 directory within the temporary directory using the `hdf5_dir`, 
but this should be saved somewhere else if the meth RSE will be reused. 

```{r, eval=TRUE}

# Create a data.frame with sample metadata associated with the array files
sample_metadata = data.frame(cancer_type = rep("COAD",  4),
  sample_type = rep("primary_tumor", 4), row.names = names(meth_array_files))

# Create meth RSE for COAD
coad_meth_rse = makeMethRSEFromArrayFiles(
  array_files = meth_array_files, 
  probe_name_column = 1, beta_value_column = 2, 
  probe_ranges = infinium_450k_probe_granges_hg19, 
  sample_metadata = sample_metadata, BPPARAM = BiocParallel::bpparam(),
  hdf5_dir = paste(tempdir(), "coad_meth_array_rse", sep = "/"))

# Display coad_meth_rse
print(coad_meth_rse)
```

## SessionInfo
```{r}
sessionInfo()
```